---
title: "RUCA = 4 Analysis"
output: html_notebook
---

```{r cache=TRUE}
library(readxl)
library(dplyr)
library(ggplot2)
#import data
 counties<- read_excel("C:/Users/peter/Downloads/2022-geography-reference-manual.xlsx")
  RUCA <- read_excel("C:/Users/peter/Downloads/ruca2010revised (1).xlsx")

#Filter only metro counties
metro_msa_codes <- counties %>%
  filter(`GEO-TYPE` =="08" & grepl("Metro Area", TITLE)) %>%
  pull(`MSA Code`) %>%
  unique()
metro_counties <- counties %>%
  filter(`GEO-TYPE` == "03", `MSA Code` %in% metro_msa_codes)
metro_counties <- metro_counties %>% 
    mutate(FIPS_State_County = paste0(`FIPS State Code`, `FIPS County Code`))
 
metro_tracts <- RUCA %>%
    semi_join(metro_counties, by = c("State-County FIPS Code" = "FIPS_State_County"))
nonmetro_tracts <- RUCA %>%
    anti_join(metro_counties, by = c("State-County FIPS Code" = "FIPS_State_County"))
nmt <- nonmetro_tracts %>%
    filter(`Primary RUCA Code 2010`==4)%>%
   filter({
    Q1 <- quantile(`Population Density (per square mile), 2010`, 0.25, na.rm = TRUE)
    Q3 <- quantile(`Population Density (per square mile), 2010`, 0.75, na.rm = TRUE)
    IQR_val <- Q3 - Q1
    `Population Density (per square mile), 2010` >= (Q1 - 2 * IQR_val) & `Population Density (per square mile), 2010` <= (Q3 + 2 * IQR_val)
  })
mt<- metro_tracts %>%
    filter(`Primary RUCA Code 2010`==4)%>%
   filter({
    Q1 <- quantile(`Population Density (per square mile), 2010`, 0.25, na.rm = TRUE)
    Q3 <- quantile(`Population Density (per square mile), 2010`, 0.75, na.rm = TRUE)
    IQR_val <- Q3 - Q1
    `Population Density (per square mile), 2010` >= (Q1 - 2 * IQR_val) & `Population Density (per square mile), 2010` <= (Q3 + 2 * IQR_val)
  })

```

Now we have all the census tracts with RUCA codes equal to four broken down by county type -- those that contain a metropolitan area, and those that do not.

```{r}
df <- tribble(
  ~Category,     ~Metric,              ~Value,
  "Metro",       "Population (millions)",         sum(mt$`Tract Population, 2010`)/1e6,
  "Nonmetro",    "Population (millions)",         sum(nmt$`Tract Population, 2010`)/1e6,
  "Metro",       "Land Area (million mi^2)",      sum(mt$`Land Area (square miles), 2010`)/1e6,
  "Nonmetro",    "Land Area (million mi^2)",      sum(nmt$`Land Area (square miles), 2010`)/1e6,
  "Metro",       "Population Density (people/mi^2)",   sum(mt$`Tract Population, 2010`)/sum(mt$`Land Area (square miles), 2010`),
  "Nonmetro",    "Population Density (people/mi^2)",  sum(nmt$`Tract Population, 2010`)/sum(nmt$`Land Area (square miles), 2010`)
)

# Plot
ggplot(df, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_fill_manual(values = c("Metro" = "darkgreen", "Nonmetro" = "lightblue")) +
  labs(title = "Metro vs Nonmetro Population Comparison (2010)",
       x = "County Type", y = NULL) +
  theme_classic() +
  theme(legend.position = "none")
sum(nmt$`Land Area (square miles), 2010`)/3018
sum(mt$`Land Area (square miles), 2010`)/1080
```

Tracts located in metropolitan counties are more dense, though tracts in non-metro counties number more and ultimately contain a greater population.

```{r}
quantile(nmt$`Tract Population, 2010`,.1)
quantile(mt$`Tract Population, 2010`,.1)
quantile(mt$`Population Density (per square mile), 2010`, .95,na.rm = TRUE)
quantile(nmt$`Population Density (per square mile), 2010`, .95,na.rm = TRUE)
sum(nmt$`Tract Population, 2010`,na.rm=TRUE)
mean(mt$`Population Density (per square mile), 2010`,na.rm = TRUE)

pull_all_tracts <- function(statistic, year) {
  library(tidycensus)
  library(tidyverse)
  
  data("fips_codes")  # Contains FIPS state and county codes
  
  # Get unique state abbreviations
  states <- unique(fips_codes$state)[1:51]  # Exclude territories
  
  # Pull data for each state and combine
  all_data <- map_dfr(states, function(st) {
    message(paste("Fetching data for", st))
    get_acs(
      geography = "tract",
      state = st,
      variables = statistic,
      year = year
    )
  })
    
  }
  
}
pull_ACS_yrs <- function(variable_code,population = "B01001_001", years, metro_trimmed, nmetro_trimmed) {
  library(tidyverse)
  library(tidycensus)
  
  for (yr in years) {
    message(paste("Pulling data for year", yr))
    
    # Pull main variable
    main_data <- pull_all_tracts(variable_code, yr) %>%
      select(GEOID, estimate) %>%
      rename(!!paste0("estimate_", yr) := estimate)
    
    # Pull population variable
    pop_data <- pull_all_tracts(population, yr) %>%
      select(GEOID, estimate) %>%
      rename(!!paste0("population_", yr) := estimate)
    
    # Merge estimates and compute rate
    merged_data <- left_join(main_data, pop_data, by = "GEOID") %>%
      mutate(!!paste0("rate_", yr) := .data[[paste0("estimate_", yr)]] / .data[[paste0("population_", yr)]])
    
    # Join with input tables
    metro_trimmed <- left_join(metro_trimmed, merged_data, by = c("State-County-Tract FIPS Code" = "GEOID"))
    nmetro_trimmed <- left_join(nmetro_trimmed, merged_data, by = c("State-County-Tract FIPS Code" = "GEOID"))
  }
  
  return(list(metro = metro_trimmed, nonmetro = nmetro_trimmed))
}
```

```{r cache=TRUE}
fedutables<-pull_ACS_yrs_fast("B14004_031","B15001_044",c(2013,2018,2023),mt,nmt)
```

```{r cache=TRUE}
fedum <- fedutables[[1]]
fedunm <- fedutables[[2]]
fedum<-read.csv("fedum")
fedunm<-read.csv("fedunm")

drop_na_estimates <- function(df) {
  df %>% filter(if_all(starts_with("estimate"), ~ !is.na(.)))
  }
fedum <-drop_na_estimates(fedum)
fedunm <- drop_na_estimates(fedunm)
```

```{r}
summarize_joined_rates <- function(metro_df, nonmetro_df, years = 2009:2023) {
  results <- tibble()
  
  for (yr in years) {
    est_col <- paste0("estimate_", yr)  # numerator
    pop_col <- paste0("population_", yr)  # denominator
    
    if (est_col %in% names(metro_df) && pop_col %in% names(metro_df)) {
      metro_rate <- sum(metro_df[[est_col]], na.rm = TRUE) / 
        sum(metro_df[[pop_col]], na.rm = TRUE)
    } else {
      metro_rate <- NA_real_
    }
    
    if (est_col %in% names(nonmetro_df) && pop_col %in% names(nonmetro_df)) {
      nonmetro_rate <- sum(nonmetro_df[[est_col]], na.rm = TRUE) / 
        sum(nonmetro_df[[pop_col]], na.rm = TRUE)
    } else {
      nonmetro_rate <- NA_real_
      }
    
      results <- bind_rows(results, tibble(
      Year = yr,
      Metro = metro_rate,
      Nonmetro = nonmetro_rate
    ))
  }
  
  return(results)
}
```

Tracts show similar outcomes in higher education enrollment rate, regardless of the presence of a metropolis in their county. Consistent with national levels, men enroll in higher education at a lower rate than women. Enrollment levels in non-metropolitan areas are slightly, but perhaps not significantly, higher than in counties containing a metropolis.

```{r}
f_enrollment<-summarize_joined_rates(fedum,fedunm,years = seq(2013,2023,by=5))
1-sum(fedunm$estimate_2023)/sum(fedunm$population_2023)
fenrollment_long <- f_enrollment %>%
  pivot_longer(cols = c(Metro, Nonmetro), names_to = "fCategory", values_to = "fEnrollmentRate")

# Plot
ggplot(fenrollment_long, aes(x = Year, y = 1-fEnrollmentRate, color = fCategory)) +
  geom_line(size = 1.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(.25, .7)) +
  scale_x_continuous(breaks = seq(2009, 2023, by = 2)) +
  scale_color_manual(values = c("Metro" = "darkgreen", "Nonmetro" = "skyblue")) +
  labs(
    title = "Proportion Enrolled in Higher Education (Females 18–24) Over Time",
    x = "Year",
    y = "Enrollment Rate",
    color = "County Type"
  ) +
  theme_minimal()


```

```{r cache=TRUE}

maleedutables<-pull_ACS_yrs_fast("B14004_015","B15001_003",c(2013,2018,2023),mt,nmt)
```

```{r cache=TRUE}
medum <- maleedutables [[1]]
medunm <- maleedutables [[2]]

medum <-drop_na_estimates(medum)
medunm <- drop_na_estimates(medunm)

m_enrollment<-summarize_joined_rates(medum,medunm,years = seq(2013,2023,by=5))

menrollment_long <- m_enrollment %>%
  pivot_longer(cols = c(Metro, Nonmetro), names_to = "mCategory", values_to = "mEnrollmentRate")

# Plot
ggplot(menrollment_long, aes(x = Year, y = 1-mEnrollmentRate, color = mCategory)) +
  geom_line(size = 1.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(.25, .7)) +
  scale_x_continuous(breaks = seq(2013, 2023, by = 2)) +
  scale_color_manual(values = c("Metro" = "darkgreen", "Nonmetro" = "skyblue")) +
  labs(
    title = "Proportion Enrolled in Higher Education (Males 18–24) Over Time",
    x = "Year",
    y = "Enrollment Rate",
    color = "County Type"
  ) +
  theme_minimal()
```

```{r}
# Prepare male data
m2enrollment_long <- m_enrollment %>%
  pivot_longer(cols = c(Metro, Nonmetro), names_to = "Category", values_to = "EnrollmentRate") %>%
  mutate(Sex = "Male", NonEnrollment =  EnrollmentRate)

# Prepare female data
f2enrollment_long <- f_enrollment %>%
  pivot_longer(cols = c(Metro, Nonmetro), names_to = "Category", values_to = "EnrollmentRate") %>%
  mutate(Sex = "Female", NonEnrollment = EnrollmentRate)

# Combine both datasets
combined_enrollment <- bind_rows(m2enrollment_long, f2enrollment_long)

# Plot
ggplot(combined_enrollment, aes(x = Year, y = 1- NonEnrollment, color = Category, linetype = Sex)) +
    geom_line(size = 1.2) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(.15, 0.5)
  ) +
  scale_x_continuous(breaks = seq(2009, 2023, by = 2)) +
  scale_color_manual(values = c("Metro" = "darkgreen", "Nonmetro" = "skyblue")) +
  labs(
    title = "Proportion Enrolled in Higher Education (Ages 18–24)",
    subtitle = "By Sex and County Type",
    x = "Year",
    y = "Enrollment Rate",
    color = "County Type",
    linetype = "Sex"
  ) +
  theme_minimal()
write.csv(fedum,file = "fedum")
write.csv(medum,file = "medum")
write.csv(fedunm,file = "fedunm")
write.csv(medunm,file = "medunm")


```

###NOW LOOK AT UNEMPLOYMENT STATISTICS

```{r}
pull_ACS_yrs("B23025_004","B23025_004",c(2009,2013,2018,2023),mt,nmt)
v<-load_variables(2013,"acs5")

```
